---
title: "Assignment #2"
format: html
---

You will need **PLINK2** installed and available in your PATH. Please follow the OS-specific setup guide in [`SETUP.md`](../../SETUP.md). The dataset for this assignment consists of the following binary PLINK files: `gwa.A2.bed`, `gwa.A2.bim`, `gwa.A2.fam` , available at the following Google Drive link: <https://drive.google.com/drive/folders/1rHoy3z52Yyj985ukjjtLhfIBxchpyYtZ?usp=drive_link>. Please download all three files and save them in `02_activities/data/`.


```{r setup, include=FALSE}
# Set up a consistent path for all chunks of code
knitr::opts_knit$set(root.dir = normalizePath("../../"))

# Force Quarto to use Git Bash for all {bash} chunks
knitr::opts_chunk$set(engine.path = list(bash = "C:/Program Files/Git/bin/bash.exe"))
```

```{r, message=FALSE, warning=FALSE, results='hide'}
library(data.table)
library(ggplot2)
library(seqminer)
library(HardyWeinberg)
library(dplyr)
library(data.table)
library(qqman)
```

```{bash}
cd ~/Downloads/gen_data/02_activities/data
pwd
head gwa.A2.fam
```

#### Question 1: Data inspection

Before you run any models, first get familiar with the dataset. You may find `data.table::fread()` in R helpful for reading `.bim` and `.fam` files.

(i) Read the `.fam` file. How many samples does the dataset contain?

```{r}
# Reading the .fam file
fam_file <- read.table("~/Downloads/gen_data/02_activities/data/gwa.A2.fam", header = FALSE)

# Count the number of samples and display the result
num_samples <- nrow(fam_file)
print(paste("Number of samples in the dataset:", num_samples))
```

(ii) Read the `.bim` file. How many SNPs does the dataset contain?

```{r}
# Read the .bim file
bim_file <- read.table("~/Downloads/gen_data/02_activities/data/gwa.A2.bim", header = FALSE)

# Count the number of SNPs and display the result
num_snps <- nrow(bim_file)
print(paste("Number of SNPs in the dataset:", num_snps))
```


(iii) Make a histogram (or density plot) of the phenotype. Does it look roughly Gaussian?

```{r}
# 1. Extracting the phenotype column (6th column of the .fam file)
phenotype <- read.table("~/Downloads/gen_data/02_activities/data/gwa.A2.fam", header = FALSE)$V6

# Filtering out missing values if present (assuming -9 indicates missing phenotypes)
phenotype_clean <- phenotype[phenotype != -9 & !is.na(phenotype)]

# 2. Make a Histogram
hist(phenotype_clean, 
     main = "Histogram of Phenotype", 
     xlab = "Phenotype Value", 
     col = "lightblue", 
     border = "black",
     breaks = 20) 

# 3. Make a Density Plot to better visualize the distribution
plot(density(phenotype_clean), 
     main = "Density Plot of Phenotype", 
     xlab = "Phenotype Value", 
     col = "darkblue", 
     lwd = 2)
```
Yes, the histogram (or density plot) displays a roughly symmetric, bell-shaped curve, indicating the phenotype follows a roughly Gaussian distribution.

#### Question 2: Quality control (QC)

Now we will perform QC using PLINK2 for the genotype files in `gwa.A2`.

(i) Using PLINK2 from the command line (bash), perform basic QC with the following filters: MAF ≥ 0.05, SNP missingness (`--geno`) ≤ 0.01, individual missingness (`--mind`) ≤ 0.10, and HWE p-value ≥ 0.00005, and output the QC’ed dataset as `gwa.qc.A2`.

```{bash}
/c/PLINK/plink2.exe --bfile ~/Downloads/gen_data/02_activities/data/gwa.A2 --maf 0.05 --geno 0.01 --mind 0.1 --hwe 0.00005 --make-bed --out ~/Downloads/gen_data/02_activities/data/gwa.qc.A2
```

#### Question 3: Relatedness

In this question, you will use **PLINK2’s built-in KING-robust kinship** (`--king-cutoff`) detect and remove related individuals.

i)  Perform LD pruning on `gwa.qc.A2` using PLINK2 with the following parameters: `--indep-pairwise 500 50 0.05`, and then generate a new dataset containing only the pruned SNPs.

```{bash}
# 1. LD pruning to identify a set of approximately independent SNPs
/c/PLINK/plink2.exe --bfile ~/Downloads/gen_data/02_activities/data/gwa.qc.A2 --indep-pairwise 500 50 0.05 --out ~/Downloads/gen_data/02_activities/data/gwa.qc.A2_pruned

# 2. Create a new dataset using only the pruned SNPs
/c/PLINK/plink2.exe --bfile ~/Downloads/gen_data/02_activities/data/gwa.qc.A2 --extract ~/Downloads/gen_data/02_activities/data/gwa.qc.A2_pruned.prune.in --make-bed --out ~/Downloads/gen_data/02_activities/data/gwa.qc.A2_pruned
```

(ii) Use PLINK2 on the LD-pruned dataset to identify a set of unrelated individuals up to (approximately) 2nd-degree relatives (use a kinship cutoff of 0.0884).

```{bash}
/c/PLINK/plink2.exe --bfile ~/Downloads/gen_data/02_activities/data/gwa.qc.A2_pruned --king-cutoff 0.0884 --out ~/Downloads/gen_data/02_activities/data/gwa_king
```

(iii) Using the unrelated individual list from part (ii), create a dataset containing only unrelated individuals and name it `gwa.qc_unrelated.A2`.

```{bash}
/c/PLINK/plink2.exe --bfile ~/Downloads/gen_data/02_activities/data/gwa.qc.A2 --keep ~/Downloads/gen_data/02_activities/data/gwa_king.king.cutoff.in.id --make-bed --out ~/Downloads/gen_data/02_activities/data/gwa.qc_unrelated.A2
```

#### Question 4: principal component analysis (PCA)

(i) Using PLINK2, perform PCA on the unrelated, LD-pruned dataset to obtain principal components.

    (Hint: Refer to the PCA section in the GWAS Tutorial II. You should use a `*.prune.in` file to select a set of independent SNPs before performing PCA.)

```{bash}
# 1. First, create a pruned dataset of unrelated individuals
/c/PLINK/plink2.exe --bfile ~/Downloads/gen_data/02_activities/data/gwa.qc_unrelated.A2 --extract ~/Downloads/gen_data/02_activities/data/gwa.qc.A2_pruned.prune.in --make-bed --out ~/Downloads/gen_data/02_activities/data/gwa.qc_unrelated_pruned

# 2. Now perform PCA:
/c/PLINK/plink2.exe --bfile ~/Downloads/gen_data/02_activities/data/gwa.qc_unrelated_pruned --pca 20 --out ~/Downloads/gen_data/02_activities/data/gwa_unrel_PCA
```

#### Question 5: GWAS analyses

We now test for association between each SNP and the continuous phenotype, adjusting for population structure using the top 3 PCs.

Assume that:

-   You will run GWAS on the unrelated, QC’ed dataset `gwa.qc_unrelated.A2`.

-   You have a covariate file (e.g., the `*.eigenvec` output from the PCA analysis) containing PC1–PC3 for each individual.

(i) Using PLINK2, run a linear regression GWAS using `gwa.qc_unrelated.A2` as the input dataset. Adjust for PCs 1–3 as covariates.

```{bash}
/c/PLINK/plink2.exe --bfile ~/Downloads/gen_data/02_activities/data/gwa.qc_unrelated.A2 --ci 0.95 --glm --covar ~/Downloads/gen_data/02_activities/data/gwa_unrel_PCA.eigenvec --covar-name PC1,PC2,PC3 --adjust --out ~/Downloads/gen_data/02_activities/data/gwa.qc_assoc_glm
```

(ii) Create a Manhattan plot of the GWAS results.

```{r}
# Load PLINK2 logistic results
assoc <- data.table::fread("~/Downloads/gen_data/02_activities/data/gwa.qc_assoc_glm.PHENO1.glm.linear")

head(assoc)

# Keep only the additive test (one row per SNP)
assoc_add <- assoc[TEST == "ADD"]

# qqman expects columns named CHR, BP, SNP, and P.
# In PLINK2 output they are #CHROM, POS, ID, and P.
setnames(assoc_add, c("#CHROM","POS","ID"), c("CHR","BP","SNP"))

# Manhattan plot
manhattan(assoc_add,
          chr = "CHR",
          bp  = "BP",
          snp = "SNP",
          p   = "P",
          xlab = "",
          ylab = "",
          suggestiveline = FALSE,
          cex.axis = 1.5,
          col = c("lightblue", "lightslateblue"))
```

(iii) Create a QQ plot of the GWAS p-values.

```{r}
# Q-Q plot
qq(assoc_add$P, main = "Q-Q plot of GWAS p-values")
```

### Criteria

| Criteria | Complete | Incomplete |
|:--------------|:----------------------------------|:----------------------|
| **Data inspection** | Correct sample and SNP counts | Counts or phenotype description/plot missing or incorrect. |
| **QC & LD pruning** | Correct PLINK2 QC command and thresholds. | QC/pruning commands, thresholds, or output datasets missing or incorrect. |
| **Relatedness & PCA** | Correct use of PLINK2 command to obtain unrelated samples and PCA run on pruned SNPs. | Relatedness step, unrelated dataset, or PCA analysis missing or incorrect. |
| **GWAS & visualisation** | Linear regression GWAS with PCs as covariates; Manhattan and QQ plots produced. | GWAS command, or Manhattan/QQ plots missing or clearly incorrect. |

### Submission Information

🚨 **Please review our [Assignment Submission Guide](https://github.com/UofT-DSI/onboarding/blob/main/onboarding_documents/submissions.md)** 🚨 for detailed instructions on how to format, branch, and submit your work. Following these guidelines is crucial for your submissions to be evaluated correctly.

------------------------------------------------------------------------

#### Submission Parameters

-   Submission Due Date: 11:59 PM – 01/04/2026

-   The branch name for your repo should be: `assignment-2`

-   What to submit for this assignment:
    -   Populate this Quarto document (`assignment_2.qmd`).
    -   Render the document with Quarto: `quarto render assignment_2.qmd`.
    -   Submit both `assignment_2.qmd`, the rendered HTML file `assignment_2.html` and saved figures in your pull request.

-   What the pull request link should look like for this assignment: `https://github.com/<your_github_username>/gen_data/pull/<pr_id>`

    -   Open a private window in your browser. Copy and paste the link to your pull request into the address bar. Make sure you can see your pull request properly. This helps the technical facilitator and learning support team review your submission easily.

------------------------------------------------------------------------

Checklist:

-   Create a branch called `assignment-2`.
-   Ensure that your repository is public.
-   Review [the PR description guidelines](https://github.com/UofT-DSI/onboarding/blob/main/onboarding_documents/submissions.md#guidelines-for-pull-request-descriptions) and adhere to them.
-   Verify that your link is accessible in a private browser window.
-   Confirm that both `assignment_2.qmd` and `assignment_2.html` are included in the pull request.
