---
title: "tutorial 2 - GWAS Analysis"
output: html
---

##### In this practical we will carry out a complete genome-wide association study (GWAS) on a small example dataset containing about 4,000 individuals and over 300K SNPs using PLINK2. PLINK2 is a widely used, open-source toolkit for GWAS and basic genetic data management. You should have **PLINK2** installed on your machine; please follow the OS-specific setup guide in [`SETUP.md`](../../SETUP.md).

### STEP 0

We will work with simulated case–control data (2,000 cases and 2,000 controls) described in the following papers:

<https://www.nature.com/articles/nprot.2010.116>

<https://www.nature.com/articles/nprot.2010.182>

Please download the three PLINK binary files: `gwa.bed`, `gwa.bim`, and `gwa.fam` from the following Google Drive link and place them in `02_activities/data/`: <https://drive.google.com/drive/folders/1EE7syaxoXmh9wKzJVMY0GfA2os3BF0Xp?usp=drive_link>

### STEP 1

#### Quality Control

In this step we perform basic QC: we remove genotyped SNPs and individuals with too much missing data, and we only keep common variants (MAF ≥ 0.05).

```{r setup, include=FALSE}
# Input files are expected in ./02_activities/data/.
#knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../../"))
library(data.table)
library(qqman)
```

```{bash,results='hide',message=FALSE, warning=FALSE}


```

Explanation of the main arguments:

-   **maf** – keeps SNPs with minor allele frequency strictly greater than 0.05. Given the small sample size, we focus on common variants.

-   **geno** – excludes SNPs with ≥1% missing genotypes.

-   **mind** – removes individuals whose genotype missingness is ≥10%.

-   **hwe** – removes markers that strongly deviate from Hardy–Weinberg equilibrium (p-value \< 5×10⁻⁵).

**OUTPUT FILES**: `gwa.qc.bed`, `gwa.qc.bim` and `gwa.qc.fam`

#### LD Pruning

We next create a subset of approximately independent SNPs. This LD-pruned set will be used for PCA and removal of related individuals.

```{bash,results='hide',message=FALSE, warning=FALSE}


```

The three numbers supplied to --indep-pairwise are:

-   Window size in SNPs (here, 500).

-   Step size (the window is shifted by 50 SNPs at a time).

-   LD threshold (r²) (0.05 in this example).

Within each window PLINK computes LD for all SNP pairs, removes one SNP from any pair with r² above the chosen threshold, then slides the window and repeats the process.

We then build a new dataset using only the pruned SNPs:

```{bash,results='hide',message=FALSE, warning=FALSE}


```


**OUTPUT FILES**: `gwa.qc_pruned.bed`, `gwa.qc_pruned.bim` and `gwa.qc_pruned.fam`

#### Detecting related or duplicate samples

Because the linear regression models used later assume that samples are independent, it is good practice to remove close relatives or duplicate individuals before running a GWAS.

PLINK2 implements the KING-robust kinship method internally. We can use the --king-cutoff option on the LD-pruned dataset to identify a set of unrelated individuals. A cutoff of 0.0884 corresponds roughly to removing up to 2nd-degree relatives.

duplicate / MZ twin: ~0.5

1st degree: ~0.25

2nd degree: ~0.125

3rd degree: ~0.0625

```{bash,results='hide',message=FALSE, warning=FALSE}


```

--king-cutoff T means: remove one individual from any pair with kinship ≥ T

This command uses the pruned SNP set to estimate pairwise kinship and then automatically chooses a subset of individuals that are mutually unrelated up to the specified cutoff. It produces ID files such as:

`./02_activities/data/gwa_qc.king.cutoff.in.id` – IDs of individuals to keep (unrelated set)

`./02_activities/data/gwa_qc.king.cutoff.out.id` – IDs of individuals to remove

We then create a dataset of unrelated samples:

```{bash,results='hide',message=FALSE, warning=FALSE}


```

#### Principal-component analysis (PCA)

To account for population structure, we compute principal components from pruned SNPs in the unrelated dataset.

We reuse the same `./02_activities/data/gwa.qc.prune.in` file to keep the same set of approximately independent SNPs:

```{bash,results='hide',message=FALSE, warning=FALSE}

```

Now perform PCA:

```{bash,results='hide',message=FALSE, warning=FALSE}


```

`./02_activities/data/gwa_unrel_PCA.eigenvec` – individual-level principal components (by default, the top 20 PCs).

`./02_activities/data/gwa_unrel_PCA.eigenval` – eigenvalues corresponding to each component.

These PCs represent major axes of genetic variation (ancestry) in the dataset. More information on PCA with PLINK can be found here: <https://speciationgenomics.github.io/pca/>

### Step 2

#### Logistic regression GWAS

Once QC, pruning, PCA, and relatedness filtering are complete, we can run the association test. We will fit a logistic regression model where:

**Phenotype (case/control) \~ SNP dosage + PC1 + PC2 + PC3**

You can later experiment with including more PCs to see how the results change.

```{bash,results='hide',message=FALSE, warning=FALSE}

```

Explanation of the main options:

-   **--logistic** : performs SNP-wise logistic regression.

-   **--covar ./02_activities/data/gwa_unrel_PCA.eigenvec** : uses the PCA output (PC scores for each individual) as a covariate file.

-   **--covar-name PC1,PC2,PC3** : includes PC1, PC2, and PC3 from that file as covariates.

-   **--ci 0.95** : requests 95% confidence intervals for effect sizes.

-   **--adjust** : produces multiple-testing adjusted p-values.

With PLINK2, the main association results are written to: `./02_activities/data/gwa.qc_assoc.PHENO1.glm.logistic`

#### Manhattan and QQ plot in R

To visualise the GWAS output, we use the qqman package to create Manhattan and Q–Q plots.

```{r}


```

```{r}


```

```{r}
# Q-Q plot

```
For more examples and plotting options, see the qqman vignette: <https://cran.r-project.org/web/packages/qqman/vignettes/qqman.html>

#### Appendix – Overview of commonly used genetic analysis tools

Below is a short description of several tools that are frequently used in GWAS and related analyses:

[PLINK]{.underline} – Classic GWAS toolkit for QC, data management, and many single-variant association tests.

[PLINK2]{.underline} – Next-generation version of PLINK with extended functionality and improved performance.

[Bcftools]{.underline} – Command-line programs for viewing, filtering, and manipulating VCF/BCF variant files.

[VCFtools]{.underline} – Suite of utilities for summarising, filtering, and converting VCF data.

[REGENIE]{.underline} – Fast whole-genome regression framework for single-variant and gene-based tests in large cohorts.

[SAIGE]{.underline} – Mixed-model method that accommodates unbalanced case–control ratios and large sample sizes.

[GCTA]{.underline} – Software for estimating SNP-heritability, genetic correlation, and performing related analyses.

[KING]{.underline} – Toolkit for inferring kinship, detecting sample duplicates, and exploring population structure.

[FastPCA]{.underline} – Efficient algorithms to compute principal components on large genotype matrices.

[STRUCTURE]{.underline} – Bayesian clustering approach for inferring population structure and assigning individuals to populations.

[ADMIXTURE]{.underline} – Fast maximum-likelihood method for estimating ancestry proportions in admixed individuals.

[ANNOVAR]{.underline} – Framework for functionally annotating genetic variants.

[GATK]{.underline} – Broad Institute toolkit for variant calling and processing of high-throughput sequencing data.

[TensorQTL]{.underline} – GPU-optimised software for rapid QTL mapping.
